Latent Variable Clustering for Classification of Events in the AT-TPC

### 2020-06-08 doc/Manuscript/article.tex
Edited section 4 -MPK

AT-TPC tracks were simulated with the \lstinline{pytpc} package developed by \citet{Bradt2017a}. Using the same parameters {\color{green} what parameters are you refering to? -MPK} as for the ${}^{46}$Ar$(p, p)$ experiment, a small set of $N=4000$ events per class was generated, as well as a larger set of $N=40000$ events per class.

@@

\subsection{Classifying  events} 
\todo{Brute force serving as motivation for why we do unsupervised learning.}
The filtering task relied on a set of $\chi^2$ distributions based on different criteria, however there was no well-defined method to quantify the effectiveness of the event selection from the thresholds applied to these distributions. In addition, the filtering task produced a binary result only, ``good" or``bad", while the data clearly contained more classes of events, if only because of the presence of carbon in the target (from the isobutane gas). In a broader perspective, an unsupervised classification algorithm would offer the possibility to ``discover" rare events not anticipated (or overlooked) by the physicist. These events would likely be filtered out using the traditional methods. From a practical point of view, compared to supervised learning, it also avoids the necessary labeling task of the learning set events, which is error prone and time consuming.

@@

The $\chi^2$ approach used in the traditional analysis performed on the $^{46}$Ar data is extremely computationally expensive because it involves the simulation of thousands of tracks for each recorded event, which are performed each iteration of the Monte-Carlo fitting sequence. Even though the reaction of interest in the $^{46}$Ar experiment had the largest cross section (elastic scattering), the sheer number of events to analyse (in the hundreds of thousands) necessitated the use of parallel processing on a high performance computing cluster. In the case of an experiment where the reaction of interest would represent less than a few percent of the total cross section, this procedure would become highly inefficient and prohibitive. Most of the CPU time would be spent on simulating events that are not of interest. An unsupervised ML algorithm able to separate the data without a priori knowledge of the different types of events increases the efficiency of the analysis tremendously, by concentrating the fitting efforts only on events of interest. In addition, the clustering allows for more exploration of the data, potentially enabling new discovery of unexpected reaction types.

@@

\begin{table}[hbtp]
\centering
\caption{Descriptions of number of events in the data. In principle we can simulate an arbitrary amount of data, but the following quantities were shown to be sufficient for these studies.{\color{blue} CITE THIS SOMEWHERE!}}\label{tab:datasets}
\begin{tabular}{lccc}
\toprule
{} & Simulated & Full & Filtered \\
\midrule
Total &  $8000$ & $51891$ & $49169$ \\
labelled & $2400$ & $1774$ &  $1582$ \\ 
\bottomrule
\end{tabular}
\end{table}

\begin{itemize}
    \item Reiterate the aim of unsupervised  clustering of events.
    \item Challenges from a machine learning perspective. 
    \begin{itemize}
        \item Supervised vs. Unsupervised learning
        \item Define Classification and Clustering 
        \item Define latent space (refer)
        \item Traditional unsupervised - distances in high dimensional spaces etc. 
    \end{itemize}
    \item Digression here to explain some ML concepts? Link back to Conv part of Michelles paper? 
    \item Opportunities presented by: 
    \begin{itemize}
        \item Transfer learning (train on VGG use on AT-TPC)
        \item Compression is understanding (leverage autoencoders) 
    \end{itemize}
    \item Building on Kuchera et. al we investigate the VGG16 networks application. 
    \item Novel contribution by applying clustering autoencoder networks. 
\end{itemize}

@@

In contrast with the classification work of \citet{Kuchera2019}, we do not have access to ground truth labels and are trying to solve a fundamentally different learning problem. Thus, rather than fine-tuning a pre-trained network under the supervised learning regime, we extract the output of the pre-trained network's last convolutional layer as a latent representation of the events on which we will apply a clustering algorithm.

@@

\begin{figure}[tb]
	\centering
	\includegraphics[width=\textwidth]{plots/mixae.pdf}
	\caption[Mixture of autoencoders schematic]{Schematic of a MIXAE model. A sample $\hat{\boldsymbol{x}}$ is compressed to set of lower-dimensional representations $\{\boldsymbol{z}^{(i)}\}$ by $N$ autoencoders. These samples are concatenated and passed through an auxiliary assignment network that predicts a confidence of cluster belonging for each autoencoder. For further details see the text. Figure adapted from \citet{Zhang}}
	\label{fig:mixae}
\end{figure}

\section{Results and Discussions}\label{sec:results}
\begin{itemize}
\item Clear out the "methods" like parts.
\item merge 5.1 and 5.2 - keeping point 1 in mind.
\item merge 5.3 5.4 and 5.6 and create table from figure 10, 11 and 12. 
\item go over 5.7 - mention instability in MIXAE results, peak performance of MIXAE is good - consistency is not. 
\item remove fig. 14 and fig 15.
\end{itemize}

>>>>

AT-TPC tracks were simulated with the \lstinline{pytpc} package developed by \citet{Bradt2017a}. Using the same parameters {\color{blue} what parameters are you referring to? -MPK} as for the ${}^{46}$Ar$(p, p)$ experiment, a small set of $N=4000$ events per class was generated, as well as a larger set of $N=40000$ events per class.

@@

\subsection{Classifying  events} 
\todo{Brute force serving as motivation for why we do unsupervised learning.}
%The filtering task relied on a set of $\chi^2$ distributions based on different criteria, 
The traditional event selection process involved a Monte-Carlo $\chi^2$ minimization method, however there was no well-defined method to quantify the effectiveness of the event selection from the thresholds applied to these distributions.
{\color{blue} I think we discussed this method earlier, but I need to verify. otherwise needs more explanation -MPK}
In addition, the selection task produced a binary result only, ``good" or``bad", relative to the event of interest, while the data clearly contained more classes of events%, if only because of the presence of carbon in the target (from the isobutane gas).
In a broader perspective, an unsupervised classification algorithm would offer the possibility to ``discover" rare events not anticipated (or overlooked) by the physicist. These events would likely be filtered out using the traditional methods. From a practical point of view, compared to supervised learning, it also avoids the necessary labeling task of the learning set events, which is error prone and time consuming.

@@

The $\chi^2$ approach used in the traditional analysis performed on the $^{46}$Ar data is extremely computationally expensive because it involves the simulation of thousands of tracks for each recorded event, which are performed each iteration of the Monte-Carlo fitting sequence. Even though the reaction of interest in the $^{46}$Ar experiment had the largest cross section (elastic scattering), the sheer number of events to analyse (in the hundreds of thousands) necessitated the use of parallel processing on a high performance computing cluster {\color{blue} is there a metric we can give here? -MPK}. In the case of an experiment where the reaction of interest would represent less than a few percent of the total cross section, this procedure would become highly inefficient and prohibitive. 
%Most of the CPU time would be spent on simulating events that are not of interest.
The computationally expensive fitting procedure would be applied to every event, instead of the few percent of the events that are of interest for the analysis.
An unsupervised ML algorithm able to separate the data without a priori knowledge of the different types of events increases the efficiency of the analysis tremendously, and allows the downstream analysis to concentrate on the fitting efforts only on events of interest. In addition, the clustering allows for more exploration of the data, potentially enabling new discovery of unexpected reaction types.

@@

\begin{table}[hbtp]
\centering
\caption{Descriptions of number of events in the data. In principle we can simulate an arbitrary amount of data, but the following quantities were shown to be sufficient for these studies.{\color{blue} reference results?}}\label{tab:datasets}
\begin{tabular}{lccc}
\toprule
{} & Simulated & Full & Filtered \\
\midrule
Total &  $8000$ & $51891$ & $49169$ \\
labelled & $2400$ & $1774$ &  $1582$ \\ 
\bottomrule
\end{tabular}
\end{table}

%\begin{itemize}
%     \item Reiterate the aim of unsupervised  clustering of events.
%     \item Challenges from a machine learning perspective. 
%     \begin{itemize}
%         \item Supervised vs. Unsupervised learning
%         \item Define Classification and Clustering 
%         \item Define latent space (refer)
%         \item Traditional unsupervised - distances in high dimensional spaces etc. 
%     \end{itemize}
%     \item Digression here to explain some ML concepts? Link back to Conv part of Michelles paper? 
%     \item Opportunities presented by: 
%     \begin{itemize}
%         \item Transfer learning (train on VGG use on AT-TPC)
%         \item Compression is understanding (leverage autoencoders) 
%     \end{itemize}
%     \item Building on Kuchera et. al we investigate the VGG16 networks application. 
%     \item Novel contribution by applying clustering autoencoder networks. 
% \end{itemize}

@@

In contrast with the classification work of \citet{Kuchera2019}, we do not assume access to ground truth labels and are trying to solve a fundamentally different learning problem. Thus, rather than fine-tuning a pre-trained network under the supervised learning regime, we extract the output of the pre-trained network's last convolutional layer as a latent representation of the events on which we will apply a clustering algorithm.

@@

\begin{figure}[tb]
	\centering
	\includegraphics[width=.8\textwidth]{plots/mixae.pdf}
	\caption[Mixture of autoencoders schematic]{Schematic of a MIXAE model. A sample $\hat{\boldsymbol{x}}$ is compressed to set of lower-dimensional representations $\{\boldsymbol{z}^{(i)}\}$ by $N$ autoencoders. These samples are concatenated and passed through an auxiliary assignment network that predicts a confidence of cluster belonging for each autoencoder. For further details see the text. Figure adapted from \citet{Zhang}}
	\label{fig:mixae}
\end{figure}
{\color{blue} This section ends quite abruptly -MPK}
\section{Results and Discussions}\label{sec:results}
\begin{itemize}
\item Clear out the "methods" like parts.
\item merge 5.1 and 5.2 - keeping point 1 in mind.
\item merge 5.3 5.4 and 5.6 and create table from figure 10, 11 and 12. 
\item go over 5.7 - mention instability in MIXAE results, peak performance of MIXAE is good - consistency is not. 
\item remove fig. 14 and fig 15.
\end{itemize}
